Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-04-05 Manuscript/manuscript.Rmd
manuscript edits

**Background.** Chromosome conformation capture sequencing technologies have shown that the three-dimensional (3D) structure of the genome is folded into distinct compartments, known as topologically associated domains (TADs) - units of coordinated gene expression. The location of TAD boundaries is highly conserved, suggesting the presence of epigenomic marks aiding in TAD formation. The ability to predict which epigenomic features are most associated with the TAD boundaries will allow to better understand the regulatory role of the 3D structure of the genome.

Existing methods for predicting associations between genomic elements tend to ignore key characteristics of genomic data. Specifically, the number of TAD boundaries is much less than the number of other genomic regions, leading to heavily imbalanced classes [??? and why this is the problem? Instead of classes, something like "leading to the model biased towards classifying the majority of the regions". Rewrite]. Furthermore, most methods utilize direct overlap as a means to quantify the association, while distance, the measure of spatial relationships, remains unaccounted for. Consequently, distances on a genomic scale vary widely, leaving uncertainty how the heavily right-tailed distribution of distance measures will affect the model’s performance.

**Methods.** We propose a novel data pre-processing pipeline that systematically addresses those shortcomings. Lasso and Elastic Net [??? Adjust methods to the current setup] were used for feature selection, coupled with five classifier methods. A number of classifier performance metrics were assessed, including the F1 measure and Matthew Correlation Coefficient (MCC), and area under the ROC curve (AUROC).

**Results.** Data preprocessing (log2-transformation and standardization) improves the performance of classification algorithms and allows for the ability to more accurately predict which genomic features are most associated with TAD boundaries. In contrast to overlap-based association measures, the distance between genomic elements was the most predictive measure. The random undersampling strategy addresses the class imbalance problem the most effectively in nearly all settings.

**Conclusions.** Current methods used to model the epigenomic features associated with TAD boundaries are insufficiently robust to handle properties of genomic data. Models applied to unprocessed data can have poor predictive performances. Focusing solely on standard performance assessment metrics, such as AUROC, can mask poor performance of the models; thus, the use of more balanced metrics, such as F1 and MCC, is warranted. Our model results in better performances and more accurate identification of important features associated with the formation of TAD boundaries.

@@

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries. Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. Although several useful insights have been made, such as the convergent binding of CTCF [@ong2014ctcf; @ghirlando2016ctcf], many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced number of boundaries vs. other regions. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, the class imbalance issue will contribute to small imbalance ratios (defined as the ratio of minority to majority classes) [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence;@jeni2013facing]. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

Commonly used measures of model performance for binary classifiers include threshold-free metrics like the area under the receiver operating curve (AUROC), and other threshold-specific metrics like accuracy, sensitivity, and specificity. The ROC curve is created by plotting the true positive rate (TPR) against the false positive rate (FPR) at various threshold settings. The subsequent AUROC, measures how well the model is capable of distinguishing between the two classes [@park2004receiver]. Accuracy is defined as the ratio of correctly identified cases (either from the positive or negative class) over the total number of cases. It measures the overall proportion of correctly classified instances. Sensitivity, also referred to as recall or TPR, is the ratio of true positives over the total number of positive classes. Specificity, on the other hand, also referred to as TNR, is the ratio of true negatives over the total number of negative classes. Each measures the proportion of correctly identified positive and negative classes respectively [@yerushalmy1947statistical]. Problems arise when using such measures in the event of imbalanced classes. For AUROC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes. Likewise, threshold-specific metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data due to being highly influenced by the correct classification of the majority class. Better measures to consider are composite metrics involving both precision and recall together because they soften the influence of a model's ability to correctly classify majority classes and yield more dependable and interpretable results. Therefore, measures like the F1-Score, Matthew's Correlation Coefficient (MCC), and balanced accuracy offer better interpretation, especially in the instance of imbalanced classes [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, balanced accuracy measures the average accuracy obtained on either class, and MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. Each metric uses all 4 quadrants of the confusion matrix, thus providing more balanced measures in the case of imbalanced data [@chicco2017ten].

@@

In this study, we have developed an ensemble framework for predicting TAD boundaries in order to systematically compare many data driven solutions used to address the issues presented above. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. Noninformative features were removed using a feature reduction algorithm and random forest classifications were performed to assess model performance, and investigate which genomic annotations were most associated with the TAD boundaries for a given system from the complete ensemble.  Model performance was evaluated using the F1-Score, MCC, and balanced accuracy.

@@

Genomic annotation data (chromatin states, histone marks, and transcription factor binding sites, GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 1).

@@

The efficacy of each re-sampling technique was compared using the F1-score, MCC, and balanced accuracy. These metrics are defined as follows:

@@

$$MCC = \frac{TP \times TN - FP \times FN}{\sqrt{ (TP + FP)(TP+FN)(TN+FP)(TN+FN) }}$$

$$Balanced \quad Accuracy = \dfrac{1}{2}\left( \dfrac{TP}{TP+FN} + \dfrac{TN}{TN+FP} \right)$$

@@

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set.

@@

We examined the class imbalance of the different data resolutions for each cell line. As expected, the total number of genomic bins decreased as resolution decreased. That is, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins (Supplementary Figure 15 (A)). The imbalance ratios, however, remained relatively stable (Supplementary Figure 15 (B)). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Ideally, a balanced dataset will have an IR close to 0.50. Here, the low imbalance ratios indicated a wide disparity between majority and minority classes (i.e. a genomic bin was much more likely to not overlap with a flanked TAD boundary than it was to overlap with one). Therefore suitable resampling techniques were needed.

@@

By isolating bins that overlapped with flanked TAD boundaries we were able to assess the correlations of genomic annotations and determine which annotations were enriched together at boundaries. To  get a hollisitic view of how some annotations tended to occur together at boundaries, we computed the average pairwise correlations between count overlaps, count percents, and $log_{2}$ distances of each genomic annotation with bins that overlapped with flanked TAD boundaries. Clustered correlograms at the 10 kb resolution, for both cell lines, showed that distinct groups of transcription factors were highly correlated including CTCF, ZNF143, RAD21, and SMC3ab (Figure 7). Likewise, combinations of particular histone modifications and chromatin states were also shown to exhibit similar profiles near TAD boundaries including H3K4me1, H3K4me2, and H3K4me3, aswell as strong and weak enhancers. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly correlated, yet inversely correlated with many transcription factor binding sites. Similar correlations were exhibited across the remainder of resolutions, for both cell lines (Supplementary Figures X-Y). The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation. The regularization was performed predictor-type and resolution specific data after implementing each of the resampling techniques within each cell line.

>>>>

**Background.** Chromosome conformation capture sequencing technologies have shown that the three-dimensional (3D) structure of the genome is folded into distinct compartments, known as topologically associated domains (TADs) - units of coordinated gene expression. The location of TAD boundaries is highly conserved, suggesting the presence of epigenomic marks aiding in TAD formation. The ability to predict which epigenomic features are most associated with the TAD boundaries will allow for better understanding the regulatory role of the 3D structure of the genome.

Existing methods for predicting associations between genomic annotations tend to ignore key characteristics of genomic data. Specifically, when constructing the model for classification, the number of genomic regions surrounding TAD boundaries is much less than the number of other regions. This leads to heavily imbalanced classes and can result in poor performances as a model's ability to assign a new sample to a given class is influenced by the prevalence of that class in the training set. Furthermore, most methods utilize direct overlap as a means to quantify this association, while distance, the measure of spatial relationships, remains unaccounted for.

**Methods.** We proposed a novel ensemble framework that systematically addresses those data irregularities and characteristics. Multiple random forest classification models, in combination with variations in cell line, data resolution, feature engineering, and re-sampling techniques were performed and compared. A number of performance metrics were assessed, including [??? update when ready].

**Results.**

**Conclusions.**

@@

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries. Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. Although several useful insights have been made, such as the convergent binding of CTCF [@ong2014ctcf; @ghirlando2016ctcf], many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced number of boundaries vs. other regions. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, the class imbalance issue will contribute to small imbalance ratios (defined as the ratio of minority to majority classes[@Orriols-Puig2008]). Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence;@jeni2013facing]. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

Commonly used measures of model performance for binary classifiers include threshold-free metrics like the area under the receiver operating curve (AUROC), and other threshold-specific metrics like accuracy, sensitivity, and specificity. The ROC curve is created by plotting the true positive rate (TPR) against the false positive rate (FPR) at various threshold settings. The subsequent AUROC, measures how well the model is capable of distinguishing between the two classes [@park2004receiver]. Accuracy is defined as the ratio of correctly identified cases (either from the positive or negative class) over the total number of cases. It measures the overall proportion of correctly classified instances. Sensitivity, also referred to as recall or TPR, is the ratio of true positives over the total number of positive classes. Specificity, on the other hand, also referred to as TNR, is the ratio of true negatives over the total number of negative classes. Each measures the proportion of correctly identified positive and negative classes respectively [@yerushalmy1947statistical]. Problems arise when using such measures in the event of imbalanced classes. For AUROC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes. Likewise, threshold-specific metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data due to being highly influenced by the correct classification of the majority class. Better measures to consider are composite metrics involving both precision and recall together because they soften the influence of a model's ability to correctly classify majority classes and yield more dependable and interpretable results. Therefore, measures like the F1-Score and Matthew's Correlation Coefficient (MCC) offer better interpretation, especially in the instance of imbalanced classes [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, while the MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. Each metric uses all 4 quadrants of the confusion matrix, thus providing more balanced measures in the case of imbalanced data [@chicco2017ten].

<!--
, and balanced accuracy offer better interpretation, especially in the instance of imbalanced classes [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, balanced accuracy measures the average accuracy obtained on either class, and MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. Each metric uses all 4 quadrants of the confusion matrix, thus providing more balanced measures in the case of imbalanced data [@chicco2017ten].
-->

@@

In this study, we have developed an ensemble framework for predicting TAD boundaries in order to systematically compare many data driven solutions used to address the issues presented above. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. Noninformative features were removed using a feature reduction algorithm and random forest classifications were performed to assess model performance, and investigate which genomic annotations were most associated with the TAD boundaries for a given system from the complete ensemble.  Model performance was evaluated using AUROC, AUPRC, F1-Score, and MCC.

@@

Genomic annotation data in the form of cell type specific chromatin states, histone marks, and transcription factor binding sites (GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 1).

@@

The efficacy of each re-sampling technique was compared using composite metrics inclusing F1-score and MCC. These metrics are defined as follows:

@@

$$MCC = \frac{TP \times TN - FP \times FN}{\sqrt{ (TP + FP)(TP+FN)(TN+FP)(TN+FN) }}$$
<!--
$$Balanced \quad Accuracy = \dfrac{1}{2}\left( \dfrac{TP}{TP+FN} + \dfrac{TN}{TN+FP} \right)$$
-->

@@

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. Additional threshold-free metrics where also reported including the AUROC and AUPRC.

@@

We examined the class imbalance of the different data resolutions for each cell line. As expected, the total number of genomic bins decreased as resolution decreased. That is, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins (Figure 4 (A)). The imbalance ratios, however, remained relatively stable (Figure 4 (B)). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Ideally, a balanced dataset will have an IR close to 0.50. Here, the low imbalance ratios indicated a wide disparity between majority and minority classes (i.e. a genomic bin was much more likely to not overlap with a flanked TAD boundary than it was to overlap with one). Therefore suitable resampling techniques were needed.

@@

By isolating bins that overlapped with flanked TAD boundaries and, in turn, counting the instances in which annotations overlapped with said bins (using the overlap counts discussed above) we were able to assess the correlations of genomic annotations and determine which annotations were enriched together at boundaries.

Clustered correlograms for known colocalized annotations showed that distinct groups of transcription factors were highly correlated including CTCF, ZNF143, RAD21, and SMC3ab (Supplementary Figures 2-9). Likewise, combinations of particular histone modifications and chromatin states were also shown to exhibit similar profiles near TAD boundaries including H3K4me1, H3K4me2, and H3K4me3, aswell as strong and weak enhancers. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly correlated, yet inversely correlated with many transcription factor binding sites. These results were consistent across resolution and cell line. The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation.
